An Extendable \proglang{Python} Implementation of Robust Optimisation Monte Carlo

### 2021-07-19 latex_files/ch3/3_2_1_general_design.tex
reordered document

\begin{figure}[H]
    \begin{center}
\includegraphics[width=0.8\textwidth]{./latex_files/graphs/ROMC.png}
    \end{center}
    \caption[Overview of the ROMC implementation.]{Overview of the
ROMC implementation. The training part follows a sequential pattern;
the functions in the green ellipses must be called in a sequential
fashion for completing the training part and define the posterior
distribution. The functions in blue ellipses are the functionalities
provided to the user.}
    \label{fig:romc_overview}
\end{figure}

@@

The prior is a uniform distribution in the range $[-2.5, 2.5]$ and the
likelihood is defined at~\ref{eq:1D_example_eq_2}. There is only one
observation $y_0 = 0$. The inference in this particular example can be
performed quite easily without using a likelihood-free inference
approach. We can exploit this fact for validating the accuracy of our
implementation. The ground-truth posterior approximated
computationally is shown in figure~\ref{fig:example_gt}.

\begin{figure}[!h]
    \begin{center}
\includegraphics[width=0.7\textwidth]{./latex_files/images/chapter3/example_gt.png}
    \end{center}
  \caption[Ground-truth posterior distribution of the simple 1D
example.]{Ground-truth posterior distribution of the simple 1D
example.}
  \label{fig:example_gt}
\end{figure}

>>>>

\begin{figure}[h]
    \begin{center}
\includegraphics[width=0.5\textwidth]{./latex_files/graphs/ROMC.png}
    \end{center}
    \caption[Overview of the ROMC implementation.]{Overview of the
ROMC implementation. The training part follows a sequential pattern;
the functions in the green ellipses must be called in a sequential
fashion for completing the training part and define the posterior
distribution. The functions in blue ellipses are the functionalities
provided to the user.}
    \label{fig:romc_overview}
\end{figure}

@@

The prior is a uniform distribution in the range $[-2.5, 2.5]$ and the
likelihood is defined at~\ref{eq:1D_example_eq_2}. There is only one
observation $y_0 = 0$. The inference in this particular example can be
performed quite easily without using a likelihood-free inference
approach. We can exploit this fact for validating the accuracy of our
implementation.

% \begin{figure}[!h]
%     \begin{center}
% \includegraphics[width=0.7\textwidth]{./latex_files/images/chapter3/example_gt.png}
%     \end{center}
%   \caption[Ground-truth posterior distribution of the simple 1D
% example.]{Ground-truth posterior distribution of the simple 1D
% example.}
%   \label{fig:example_gt}
% \end{figure}

### 2021-07-19 latex_files/ch3/3_2_2_training.tex
reordered document

\begin{Code}
---------------------------------- python ----------------------------------
>>> romc.fit_posterior(n1,
                       eps_filter,
                       use_bo=False,
                       optimizer_args=None,
                       seed=None,
                       use_surrogate=None,
                       region_args=None,
                       fit_models=False,
                       fit_models_args=None,
                       eps_region=None,
                       eps_cutoff=None)
----------------------------------------------------------------------------
\end{Code}

\noindent
The function \code{fit_posterior} is a syntactic sugar for applying
\code{solve_problems} and \code{estimate_regions} into a single
step.

% merges all steps for constructing the bounding box into
% a single command. If the user doesn't want to manually inspect the
% histogram of the distances before deciding where to set the threshold
% $\epsilon$, he may call \code{romc.fit_posterior()} and the whole
% training part will be done end-to-end. There are two alternatives for
% setting the threshold $\epsilon$; the first is to set to a specific
% value blindly, and the second is to set at as a specific quantile of
% the histogram of distances. In the second scenario the
% \code{quantile} argument must be set to a floating number in the
% range $[0,1]$ and \code{eps='auto'}.

\begin{Code}
---------------------------------- python ----------------------------------
>>> romc.visualize_region(i)
----------------------------------------------------------------------------
\end{Code}

\noindent

It can be used for plotting the bounding box around the optimal point,
when the parameter space is $1D$ or $2D$. The argument \code{i} is the
index of the corresponding optimization problem i.e.\ $i<n_1$.

\begin{Code}
---------------------------------- python ----------------------------------  
>>> romc.compute_eps(quantile)
----------------------------------------------------------------------------
\end{Code}

\noindent
This function return the appropriate distance value $d_{i=\kappa}^*$
where $\kappa = \lfloor \frac{quantile}{n} \rfloor$ from the
collection $\{ d_i^* \} \forall i = \{1, \ldots, n\}$ where $n$ is the
number of accepted solutions. It can be used to automate the selection
of the threshold $\epsilon$, e.g.\
\code{eps=romc.compute_eps(quantile=0.9)}.

@@

\begin{Code}
------------------------------ python snippet ------------------------------  
  n1 = 500 # number of optimisation problems
  seed = 21 # seed for solving the optimisation problems
  eps = .75 
  use_bo = False # set to True for switching to Bayesian optimisation

@@

\begin{figure}[h]
  \begin{center}
    \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter3/example_theta_dist.png}
    \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter3/example_region.png}
    \end{center}
    \caption[Histogram of distances at the 1D example.]{Histogram of
      distances and visualisation of a specific region.}
    \label{fig:example_training_hist}
\end{figure}

>>>>

\begin{Code}
---------------------------------- python ----------------------------------
>>> romc.fit_posterior(args*)  # training in a single call
>>> romc.visualize_region(i)   # acceptance region
>>> romc.compute_eps(quantile) # estimates ε
----------------------------------------------------------------------------
\end{Code}

\noindent
The function \code{fit_posterior} is a syntactic sugar for applying
\code{solve_problems} and \\ \code{estimate_regions} into a single
step. The function \code{visualize_region} can be used for plotting
the bounding box around the optimal point, when the parameter space is
$1D$ or $2D$. The argument \code{i} is the index of the corresponding
optimization problem i.e.\ $i<n_1$. \code{compute_eps} returns the
appropriate distance value $d_{i=\kappa}^*$ where
$\kappa = \lfloor \frac{quantile}{n} \rfloor$ from the collection
$\{ d_i^* \} \forall i = \{1, \ldots, n\}$ where $n$ is the number of
accepted solutions. It can be used to automate the selection of the
threshold $\epsilon$, e.g.\ \code{eps=romc.compute_eps(quantile=0.9)}.

@@

\begin{Code}
------------------------------ python snippet ------------------------------
  # Trainining (fitting) part
  n1 = 500 # number of optimisation problems
  seed = 21 # seed for solving the optimisation problems
  eps = .75 
  use_bo = False # set to True for switching to Bayesian optimisation

@@

% \begin{figure}[h]
%   \begin{center}
%     \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter3/example_theta_dist.png}
%     \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter3/example_region_samples.png}
%     \end{center}
%     \caption[Histogram of distances at the 1D example.]{Histogram of
%       distances and visualisation of a specific region.}
%     \label{fig:example_training_hist}
% \end{figure}

\begin{figure}[h]
  \begin{center}
    \includegraphics[width=0.32\textwidth]{./latex_files/images/chapter3/example_theta_dist.png}
    \includegraphics[width=0.32\textwidth]{./latex_files/images/chapter3/example_region_samples.png}
    \includegraphics[width=0.32\textwidth]{./latex_files/images/chapter3/example_posterior.png}

    \end{center}
    \caption[Histogram of distances at the 1D example.]{Histogram of
      distances and visualisation of a specific region.}
    \label{fig:example_training_hist}
\end{figure}

### 2021-07-19 latex_files/ch3/3_2_3_inference.tex
reordered document

\begin{Code}
---------------------------------- python ----------------------------------
>>> romc.compute_expectation(h)
----------------------------------------------------------------------------
\end{Code}

\noindent
This function computes the expectation
$E_{p(\thetab|\data)}[h(\thetab)]$ using
expression~\eqref{eq:expectation}. The argument \code{h} can be
any python \code{Callable}.

\begin{Code}
---------------------------------- python ----------------------------------
>>> romc.eval_unorm_posterior(theta, eps_cutoff=False)
----------------------------------------------------------------------------
\end{Code}

\noindent
This function computes the unnormalised posterior approximation using
expression~\eqref{eq:approx_posterior}.

\begin{Code}
---------------------------------- python ----------------------------------
>>> romc.eval_posterior(theta, eps_cutoff=False)
----------------------------------------------------------------------------
\end{Code}

\noindent
This function evaluates the normalised posterior estimating the
partition function $Z = \int p_{d,\epsilon}(\thetab|\data)d\thetab$;
with Riemann's integral approximation. The approximation is
computationally feasible only in low-dimensional parametric
spaces.

@@

\begin{Code}
------------------------------ python snippet ------------------------------  
  seed = 21
  n2 = 50
  romc.sample(n2=n2, seed=seed)

@@

  # compute expectation
  print("Expected value   : %.3f" % romc.compute_expectation(h=lambda x: np.squeeze(x)))
  # Expected value   : -0.079

  print("Expected variance: %.3f" % romc.compute_expectation(h=lambda x: np.squeeze(x)**2))
  # Expected variance: 1.061

@@

\begin{figure}[h]
    \begin{center}
      \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter3/example_marginal.png}
      \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter3/example_region_samples.png}
    \end{center}
  \caption[Histogram of the obtained samples at the 1D example.]{(a) Left: Histogram of the obtained samples. (b) Right: Acceptance region around $\theta_1^*$ with the obtained samples plotted inside.}
  \label{fig:example_sampling}
\end{figure}

\begin{figure}[h]
    \begin{center}
      \includegraphics[width=0.75\textwidth]{./latex_files/images/chapter3/example_posterior.png}
    \end{center}
  \caption[Approximate posterior evaluation, at the 1D example.]{Approximate posterior evaluation.}
  \label{fig:approx_posterior}
\end{figure}

>>>>

\begin{Code}
---------------------------------- python ----------------------------------
>>> romc.compute_expectation(h) # E[h(x)|theta]
>>> romc.eval_unorm_posterior(theta, eps_cutoff=False) # unnorm p(theta)
>>> romc.eval_posterior(theta, eps_cutoff=False) # normalised p(theta)
----------------------------------------------------------------------------
\end{Code}

\noindent
This function computes the expectation
$E_{p(\thetab|\data)}[h(\thetab)]$ using
expression~\eqref{eq:expectation}. The argument \code{h} can be any
python \code{Callable}. \code{eval_unnorm_posterior} computes the
unnormalised posterior approximation using
expression~\eqref{eq:approx_posterior}. \code{eval_posterior}
evaluates the normalised posterior estimating the partition function
$Z = \int p_{d,\epsilon}(\thetab|\data)d\thetab$; with Riemann's
integral approximation. The approximation is computationally feasible
only in low-dimensional parametric spaces.

@@

\begin{Code}
------------------------------ python snippet ------------------------------  
  # Inference part
  seed = 21
  n2 = 50
  romc.sample(n2=n2, seed=seed)

@@

  # compute expectation
  exp_val = romc.compute_expectation(h=lambda x: np.squeeze(x))
  print("Expected value   : %.3f" % exp_val)
  # Expected value   : -0.079

  exp_var = romc.compute_expectation(h=lambda x: np.squeeze(x)**2)
  print("Expected variance: %.3f" % exp_var)
  # Expected variance: 1.061

@@

% \begin{figure}[h]
%     \begin{center}
%       \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter3/example_marginal.png}
%     \end{center}
%   \caption[Histogram of the obtained samples at the 1D example.]{(a) Left: Histogram of the obtained samples. (b) Right: Acceptance region around $\theta_1^*$ with the obtained samples plotted inside.}
%   \label{fig:example_sampling}
% \end{figure}

% \begin{figure}[h]
%     \begin{center}
%       \includegraphics[width=0.6\textwidth]{./latex_files/images/chapter3/example_posterior.png}
%     \end{center}
%   \caption[Approximate posterior evaluation, at the 1D example.]{Approximate posterior evaluation.}
%   \label{fig:approx_posterior}
% \end{figure}

### 2021-07-19 latex_files/ch3/3_2_4_evaluation.tex
reordered document

\begin{Code}
---------------------------------- python ----------------------------------  
>>> romc.compute_divergence(gt_posterior,
                            bounds=None,
                            step=0.1,
                            distance="Jensen-Shannon")
----------------------------------------------------------------------------
\end{Code}

\noindent
This function computes the divergence between the ROMC approximation
and the ground truth posterior. Since the computation is performed
using the Riemann's approximation, this method can only work in low
dimensional parameter spaces; it is suggested to be used for up to the
three-dimensional parameter space. As mentioned at the beginning of
this chapter, in a real-case scenario it is not expected the
ground-truth posterior to be available. However, in cases where the
posterior can be approximated decently well with a numerical approach
(as in the current example) or with some other inference approach,
this function can provide a numerical measure of the agreement between
the two approximations. The argument \code{step} defines the step used
in the Riemann's approximation and the argument \code{distance} can
take either the \code{Jensen-Shannon} or the \code{KL-divergence}
value, for computing the appropriate distance.

\begin{Code}
---------------------------------- python ----------------------------------  
>>> romc.compute_ess()
----------------------------------------------------------------------------
\end{Code}

\noindent
This function computes the Effective Sample Size (ESS) using the
following expression~\cite{Sudman1967},

@@

\begin{Code}
------------------------------ python snippet ------------------------------  
  res = romc.compute_divergence(wrapper, distance="Jensen-Shannon")                                 
  print("Jensen-Shannon divergence: %.3f" % res)
  # Jensen-Shannon divergence: 0.025

>>>>

\begin{Code}
---------------------------------- python ----------------------------------  
>>> romc.compute_divergence(gt_posterior,
                            bounds=None,
                            step=0.1,
                            distance="Jensen-Shannon")
>>> romc.compute_ess()
----------------------------------------------------------------------------
\end{Code}

\noindent
This function computes the divergence between the ROMC approximation
and the ground truth posterior. Since the computation is performed
using the Riemann's approximation, this method can only work in low
dimensional parameter spaces; it is suggested to be used for up to the
three-dimensional parameter space. As mentioned at the beginning of
this chapter, in a real-case scenario it is not expected the
ground-truth posterior to be available. However, in cases where the
posterior can be approximated decently well with a numerical approach
(as in the current example) or with some other inference approach,
this function can provide a numerical measure of the agreement between
the two approximations. The argument \code{step} defines the step used
in the Riemann's approximation and the argument \code{distance} can
take either the \code{Jensen-Shannon} or the \code{KL-divergence}
value, for computing the appropriate distance. \code{compute_ess}
computes the Effective Sample Size (ESS) using the following
expression~\cite{Sudman1967},

@@

\begin{Code}
------------------------------ python snippet ------------------------------  
  # Evaluation part
  res = romc.compute_divergence(wrapper, distance="Jensen-Shannon")                                 
  print("Jensen-Shannon divergence: %.3f" % res)
  # Jensen-Shannon divergence: 0.025

### 2021-07-19 latex_files/ch4/4_2_ma2.tex
reordered document

\begin{figure}[H]
    \begin{center}
      \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter4/ma2_region_3.png}
      \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter4/ma2_region_3_bo.png}
    \end{center}
  \caption[The acceptance region of a specific deterministic simulator.]{The acceptance region in a specific optimisation problem. In the left figure the region obtained with gradient-based optimiser and in the right one with Bayesia Optimisation.}
  \label{fig:ma2_5}
\end{figure}

\begin{figure}[H]
    \begin{center}
      \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter4/mae2_hist_t1_rejection.png}
      \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter4/mae2_hist_t2_rejection.png}\\
      \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter4/mae2_hist_t1_romc.png}
      \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter4/mae2_hist_t2_romc.png}\\
      \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter4/mae2_hist_t1_romc_bo.png}
      \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter4/mae2_hist_t2_romc_bo.png}\\
    \end{center}
    \caption[MA2 example, evaluation of the marginal
    distributions.]{Histogram of the marginal posterior distributions
      using three different inference approaches; (a) in the first
      row, the samples are obtained using Rejection ABC sampling (b)
      in the second row, using ROMC with a gradient-based optimiser
      and (c) in the third row, using ROMC with Bayesian optimization
      approach. The vertical (red) line represents the samples mean
      $\mu$ and the horizontal (black) the standard deviation
      $\sigma$.}
  \label{fig:ma2_3}
\end{figure}

>>>>

\begin{figure}[h]
    \begin{center}
      \includegraphics[width=0.35\textwidth]{./latex_files/images/chapter4/ma2_region_3.png}
      \includegraphics[width=0.35\textwidth]{./latex_files/images/chapter4/ma2_region_3_bo.png}
    \end{center}
  \caption[The acceptance region of a specific deterministic simulator.]{The acceptance region in a specific optimisation problem. In the left figure the region obtained with gradient-based optimiser and in the right one with Bayesia Optimisation.}
  \label{fig:ma2_5}
\end{figure}

\begin{figure}[h]
    \begin{center}
      \includegraphics[width=0.35\textwidth]{./latex_files/images/chapter4/mae2_hist_t1_rejection.png}
      \includegraphics[width=0.35\textwidth]{./latex_files/images/chapter4/mae2_hist_t2_rejection.png}\\
      \includegraphics[width=0.35\textwidth]{./latex_files/images/chapter4/mae2_hist_t1_romc.png}
      \includegraphics[width=0.35\textwidth]{./latex_files/images/chapter4/mae2_hist_t2_romc.png}\\
      \includegraphics[width=0.35\textwidth]{./latex_files/images/chapter4/mae2_hist_t1_romc_bo.png}
      \includegraphics[width=0.35\textwidth]{./latex_files/images/chapter4/mae2_hist_t2_romc_bo.png}\\
    \end{center}
    \caption[MA2 example, evaluation of the marginal
    distributions.]{Histogram of the marginal posterior distributions
      using three different inference approaches; (a) in the first
      row, the samples are obtained using Rejection ABC sampling (b)
      in the second row, using ROMC with a gradient-based optimiser
      and (c) in the third row, using ROMC with Bayesian optimization
      approach. The vertical (red) line represents the samples mean
      $\mu$ and the horizontal (black) the standard deviation
      $\sigma$.}
  \label{fig:ma2_3}
\end{figure}
